---
title: "cambridgema vs ggplot2 Default Colors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cambridgema vs ggplot2 Default Colors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
library(cambridgema)
library(ggplot2)
```

## Side-by-Side Comparisons

This vignette demonstrates the visual differences between ggplot2's default
color scales and cambridgema palettes across common plot types.

### Scatter Plot: Categorical Colors

**ggplot2 Default**

```{r scatter-default, fig.height=4}
ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "ggplot2 Default Colors") +
  theme_minimal()
```

**cambridgema autumn**

```{r scatter-cam, fig.height=4}
ggplot(iris, aes(Sepal.Width, Sepal.Length, color = Species)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_cam("autumn") +
  labs(title = "cambridgema: autumn palette") +
  theme_minimal()
```

### Bar Chart: Fill Colors

**ggplot2 Default**

```{r bar-default, fig.height=4}
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  labs(title = "ggplot2 Default Colors") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

**cambridgema summer**

```{r bar-cam, fig.height=4}
ggplot(mpg, aes(class, fill = class)) +
  geom_bar() +
  scale_fill_cam("summer") +
  labs(title = "cambridgema: summer palette") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

### Heatmap: Sequential Colors

**ggplot2 Default**

```{r heat-default, fig.height=4}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  labs(title = "ggplot2 Default Sequential") +
  theme_minimal()
```

**cambridgema river_seq**

```{r heat-cam, fig.height=4}
ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_tile() +
  scale_fill_cam("river_seq", discrete = FALSE) +
  labs(title = "cambridgema: river_seq palette") +
  theme_minimal()
```

### Correlation Matrix: Diverging Colors

**ggplot2 Default**

```{r cor-default, fig.height=5}
cor_mat <- cor(mtcars[, 1:6])
cor_df <- as.data.frame(as.table(cor_mat))
names(cor_df) <- c("Var1", "Var2", "value")

ggplot(cor_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  labs(title = "ggplot2 Default") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**cambridgema crimson_blue**

```{r cor-cam, fig.height=5}
ggplot(cor_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  scale_fill_cam("crimson_blue") +
  labs(title = "cambridgema: crimson_blue palette") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Box Plot: Group Comparisons

**ggplot2 Default**

```{r box-default, fig.height=4}
ggplot(iris, aes(Species, Sepal.Length, fill = Species)) +
  geom_boxplot() +
  labs(title = "ggplot2 Default Colors") +
  theme_minimal() +
  theme(legend.position = "none")
```

**cambridgema spring**

```{r box-cam, fig.height=4}
ggplot(iris, aes(Species, Sepal.Length, fill = Species)) +
  geom_boxplot() +
  scale_fill_cam("spring") +
  labs(title = "cambridgema: spring palette") +
  theme_minimal() +
  theme(legend.position = "none")
```

## Palette Overview

View all available palettes:

```{r demo, fig.height=8}
plot_cam_demo()
```

## Choosing a Palette

| Data Type | Recommended Palettes |
|-----------|---------------------|
| Categorical (3-5 groups) | `spring`, `summer`, `autumn`, `winter`, `river` |
| Sequential (continuous) | `river_seq`, `crimson_seq`, `foliage_seq` |
| Diverging (centered on zero) | `crimson_blue`, `brick_pine` |

### Quick Reference

```{r list}
list_palettes()
```

## Combining Color with Shape

For maximum clarity, pair color with other visual encodings:

```{r shapes, fig.height=4}
ggplot(iris, aes(Sepal.Width, Sepal.Length,
                 color = Species, shape = Species)) +
  geom_point(size = 3) +
  scale_color_cam("autumn") +
  labs(title = "Color + Shape Encoding") +
  theme_minimal()
```

## Accessibility Tools

Check color distinguishability for any palette:

```{r check}
check_colorblind("autumn")
```

Visualize how a palette appears under different viewing conditions:

```{r sim, fig.height=6}
plot_colorblind_sim("autumn")
```
